---
title: "spatialGE: Basic tutorial for visualization and spatial heterogeneity analysis"
output: rmarkdown::melanoma_ST_vignette
vignette: >
  %\VignetteIndexEntry{spatialGE_tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The package spatialGE provides a set of tools for the visualization of gene expression from 
spatially-resolved arrays, such as those generated by the 10X Visium platform. 

# Installation
The spatialGE repository is available at GitHub and can be installed via `devtools`.
To install `devtools` in case is not installed in your R console, please run the 
following code:
```{r}
if("devtools" %in% rownames(installed.packages()) == FALSE) {
  install.packages("devtools")
}
```

After making sure `devtolls` is installed, proceed to install `spatialGE`:
```{r}
devtools::install_github("fridleylab/spatialGEdev")
```

To use `spatialGE`, load the package using the command:
```{r setup}
library(spatialGEdev)
```

# Spatially-resolved expression of melanoma stage III tumor biopsies

